The Chemistry Development Kit (CDK): atom typing, rendering, molecular formulas, and substructure searching

### 2016-10-09 article.tex
More consistent with the rest of the paper
More high impact metabolomics work using the CDK

The open source cheminformatics community has made significant steps
forward recently~\cite{OBoyle2011b} as evidenced by the growing number
of tools and underlying toolkits, along with the usage of these
software components in a variety of applications.
The Chemistry Development Kit (CDK) is one of the tools developed
under the aegis of the Blue Obelisk 
and previously documented versions have been widely adopted~\cite{Steinbeck2003,Steinbeck2006}.
Use of the CDK ranges from inclusion of CDK functionality in
wrapper platforms such as Cinfony~\cite{OBoyle2008}, incorporation
within the R environment (rcdk~\cite{Guha2007}), and as plugins for
Taverna~\cite{Truszkowski2011}, 
KNIME~\cite{Beisken2013}, Cytoscape (ChemViz2~\cite{ChemViz2}), and for
Microsoft Excel (LICSS~\cite{Lawson2012}).
In contrast to scenarios that have made CDK functionality available in
larger systems, a number of projects have employed the CDK as a
general cheminformatics toolkit. Examples include 
jCompoundMapper~\cite{Hinselmann2011}, ScaffoldHunter~\cite{wetzel2009interactive}, OMG~\cite{Peironcely2012},
PaDEL~\cite{yap2011padel}, ReactPRED~\cite{ReactPRED}, SMSD~\cite{Rahman2009,Rahman2014,Rahman2016},
WhichCyp~\cite{Rostkowski2013}, MetaPrint2D~\cite{Carlsson2010}, and the IUPHAR/BPS Guide to Pharmacology database~\cite{Southan2016}.
A number of such tools were initially developed using older versions
of the CDK and are updated to new releases as they are made
available. Examples include Bioclipse~\cite{spjuth2007bioclipse,
spjuth2009bioclipse} and
AMBIT~\cite{jeliazkova2011ambit,jeliazkova2011ambitsmarts,kochev2013ambit}. The
CDK has also played a role in a number of chemical studies, such as finding
the maximally bridging rings in chemical structures~\cite{Marth2015},
prediction of organic reactions~\cite{Segler2016}, and bioactivities
of compounds~\cite{Alvarsson2016}.

@@

  While the API itself has not been significantly changed, the
  internals have been revamped. In addition to improved overlap
  resolution noted above, the engine appropriately handles large ring
  systems and maintains input stereochemistry. While previous CDK
versions supported 
  double bond stereochemistry the new engine is more efficient in
  using this information when generate 2D layouts. Furthermore, the
  engine assigns wedge bond information based on tetrahedral
  stereochemistry. These features are exemplified by the following
  code and the resulting layout depiction in
  Figure~\ref{fig:sdgstereo}.

@@

A chemical formula is the simplest chemical representation of a
compound.  It defines the number of isotopes or elements that
compose  a compound without describing how atoms
are bonded. With the rise of metabolomics it has become increasingly
relevant to have full support for these in cheminformatics
libraries~\cite{RojasCherto2011}.

>>>>

The open source cheminformatics community has made significant steps
forward recently~\cite{OBoyle2011b} as evidenced by the growing number
of tools and underlying toolkits, along with the usage of these
software components in a variety of applications.
The Chemistry Development Kit (CDK) is one of the tools developed
under the aegis of the Blue Obelisk 
and previously documented versions have been widely adopted~\cite{Steinbeck2003,Steinbeck2006}.
Use of the CDK ranges from inclusion of CDK functionality in
wrapper platforms such as Cinfony~\cite{OBoyle2008}, incorporation
within the R environment (rcdk~\cite{Guha2007}), and as plugins for
Taverna~\cite{Truszkowski2011}, 
KNIME~\cite{Beisken2013}, Cytoscape (ChemViz2~\cite{ChemViz2}), and for
Microsoft Excel (LICSS~\cite{Lawson2012}).
In contrast to scenarios that have made CDK functionality available in
larger systems, a number of projects have employed the CDK as a
general cheminformatics toolkit. Examples include 
jCompoundMapper~\cite{Hinselmann2011}, ScaffoldHunter~\cite{wetzel2009interactive}, OMG~\cite{Peironcely2012},
PaDEL~\cite{yap2011padel}, ReactPRED~\cite{ReactPRED}, SMSD~\cite{Rahman2009,Rahman2014,Rahman2016},
WhichCyp~\cite{Rostkowski2013}, MetaPrint2D~\cite{Carlsson2010}, MetFrag~\cite{Wolf2010},
and the IUPHAR/BPS Guide to Pharmacology database~\cite{Southan2016}.
A number of such tools were initially developed using older versions
of the CDK and are updated to new releases as they are made
available. Examples include Bioclipse~\cite{spjuth2007bioclipse,
spjuth2009bioclipse} and
AMBIT~\cite{jeliazkova2011ambit,jeliazkova2011ambitsmarts,kochev2013ambit}. The
CDK has also played a role in a number of chemical studies, such as finding
the maximally bridging rings in chemical structures~\cite{Marth2015},
prediction of organic reactions~\cite{Segler2016}, and bioactivities
of compounds~\cite{Alvarsson2016}.

@@

  While the API itself has not been significantly changed, the
  internals have been revamped. In addition to improved overlap
  resolution noted above, the engine appropriately handles large ring
  systems and maintains input stereochemistry. While previous CDK
versions supported 
  double bond stereochemistry the new engine is more efficient in
  using this information when generate 2D layouts. Furthermore, the
  engine assigns wedge bond information based on tetrahedral
  stereochemistry. These features are exemplified by the following
  code and the resulting layout depiction in
  Figure~\ref{fig:sdgstereo}:

@@

A chemical formula is the simplest chemical representation of a
compound.  It defines the number of isotopes or elements that
compose  a compound without describing how atoms
are bonded. With the rise of metabolomics it has become increasingly
relevant to have full support for these in cheminformatics
libraries~\cite{Wolf2010,RojasCherto2011,Pluskal2012,Pluskal2010}.
